---
title: "Training model classifying a cell subtype from scRNA-seq data"
author: "Vy Nguyen"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{3. Training child model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

## Introduction

Apart from a basic model for a basic independent cell type, our methods also 
supports cell types considered as the children 
of another cell type. In this case, the so-called parent cell type must already
be represented by a classification model.

Here we consider the model classifying child cell type as child model and of
course, parent model is used for classification of parent cell type.

Child model is used to distinguish a particular child cell type from other 
children cell types of the same parent. Therefore, our methods will examine 
all cells by parent model before training and testing for child model. Only 
cells that are considered as parent cell type will be used to train and 
test the new model.

The basis for this approach is that while markers that are very well suited
to differentiate a CD4+ T cell from a CD8+ T cell, these markers may worsen 
the differentiation of a T cell *vs.* other cells.

## Parent model
A first prerequisite of training for a child model is the parent model. 
A parent model is of class scTypeR and must be available in the 
working space, among default pretrained models of the package or among 
trained models in a user supplied database.

In this example, we load B cells classifier in the package default 
models to our working place.
```{r}
# use BiocManager to install from Bioconductor
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

# the scTypeR package
if (!require(scTypeR))
  BiocManager::install("scTypeR")
```

```{r}
data("default_models")
clf_B <- default_models[['B cells']]
clf_B
```
## Preparing train object and test object

Same as training for basic models, training for a child model also requires 
a train (Seurat/SCE) object and a test (Seurat/SCE) object. All objects must 
have a slot in meta data indicating the type of cells. Tag slot indicating 
parent cell type can also be provided. In this case, parent cell type tag 
will further be tested for coherence with the provided parent classifier.

Cell tagged as child cell type but incoherent to parent cell type will be 
removed from training and testing for the child cell type classifier. 

In this vignette, we use the human lung dataset from Zilionis et al., 2019,
which is available in the scRNAseq (2.4.0) library. The dataset is stored as a
SCE object.

To start the training workflow, we first load the neccessary libraries. 
```{r}
# we use the scRNAseq package to load example data
if (!require(scRNAseq))
  BiocManager::install("scRNAseq")
```

First, we load the dataset. To reduce the computational complexity of this vignette, we only use the first 5000 cells of the dataset.

```{r}
zilionis <- ZilionisLungData()
zilionis <- zilionis[, 1:5000]
```

We split this dataset into two parts, one for the training and the other for the testing.

```{r}
pivot = ncol(zilionis)%/%2
train_set <- zilionis[, 1:pivot]
test_set <- zilionis[, (1+pivot):ncol(zilionis)]
```


In this dataset, the cell type meta data is stored in the *Most likely LM22 cell type*
slot of the SingleCellExperiment object (in both the train object and test object). 

If the cell type is stored not stored as the default identification (set through
`Idents` for Seurat object) the slot must be set as a parameter in the training
and testing function (see below).

```{r}
table(train_set$`Most likely LM22 cell type`)
```
```{r}
table(test_set$`Most likely LM22 cell type`)
```

Unlike the example of the training basic model, we will remove all NAs cells in order to reduce the computational complexity for this example.

```{r}
# remove NAs cells
train_set <- train_set[, !is.na(train_set$`Most likely LM22 cell type`)]
test_set <- test_set[, !is.na(test_set$`Most likely LM22 cell type`)]
```

```{r}
# convert cell label: 
# 1 - positive to plasma cells, 
# 0 - negative to plasma cells
train_set$plasma <- unlist(lapply(train_set$`Most likely LM22 cell type`,
                                  function(x) if (x == 'Plasma cells') {1} else {0}))

test_set$plasma <- unlist(lapply(test_set$`Most likely LM22 cell type`,
                                 function(x) if (x == 'Plasma cells') {1} else {0}))
```

We may want to check the number of cells in each category:
```{r}
table(train_set$plasma)
# 1: plasma cells, 0: not plasma cells
```

## Defining set of features

Next, we define a set of features, which will be used for the child 
classification model. Supposing we are training a model for classifying plasma 
cells, we define the set of features as follows:

```{r}
selected_features_plasma <- c("SDC1", "CD19", "BACH2", "CD74", 'CD27', "CD37", 
                              'IGHG1', 'IGHG2', 'IGHM', "CD40LG", "MCL1", "MZB1")
```

## Train model

Training for a child model needs more parameters than training a basic one.
Users must indicate the parent classifier. There are three ways to indicate the
parent classifier to the train method:
  
  * Users can use an available model in current working place.
  
  * Users can give name of a model among default pretrained models, 
for example: *parent_cell = 'B cells'*
  
  * Users can give name of a model among models available in users' database 
AND the path to that database, for example: `parent_cell = 'B cells', path.to.models = '.'`

Train the child classifier:
```{r}
set.seed(123)
clf_plasma <- train_classifier(train_obj = train_set, 
features = selected_features_plasma, cell_type = "Plasma cells", 
sce_assay = 'counts', sce_tag_slot = 'plasma', parent_clf = clf_B)
```
If the  cells classifier has not been loaded to the current working space, 
an equivalent training process should be:
```{r}
set.seed(123)
clf_plasma <- train_classifier(train_obj = train_set, 
features = selected_features_plasma, cell_type = "Plasma cells", 
sce_assay = 'counts', sce_tag_slot = 'plasma', parent_cell = 'B cells')
```
```{r}
clf_plasma
```
```{r}
clf(clf_plasma)
```

## Test model

The parent classifier must be also set in test method.
```{r}
clf_plasma_test <- test_classifier(test_obj = test_set, 
classifier = clf_plasma, sce_assay = 'counts', sce_tag_slot = 'plasma', 
parent_clf = clf_B)
```

### Interpreting test model result
The test result obtained from a child model can be interpreted in the same way
as we do with the model for basic cell types. We can change the prediction 
probability threshold according to the research project or personal preference 
and plot a roc curve.

```{r}
print(clf_plasma_test$auc)
roc_curve <- plot_roc_curve(test_result = clf_plasma_test)
plot(roc_curve)
```

## Save child classification model for further use

In order to save child classifier, the parent classifier must already exist in the 
classifier database, either in the package default database or in the user-defined 
database.

```{r}
# see list of available model in package
data("default_models")
names(default_models)
```

In our package, the default models already include a model classifying plasma cells.
Therefore, we will save this model to a new local database specified by the 
*path.to.models* parameter. If you start with a fresh new local database, 
there is no available parent classifier of plasma cells' classifier. Therefore,
we have to save the parent classifier first, e.g. the classifier for B cells.

```{r}
# no copy of pretrained models is performed
save_new_model(new_model = clf_B, path.to.models = getwd(), 
               include.default = FALSE) 
save_new_model(new_model = clf_plasma, path.to.models = getwd(), 
               include.default = FALSE) 
```

## Applying newly trained models for cell classification

When we save the B cells' classifier and the plasma cells' classifier, a local database is newly created. We can use this new database to classify cells in a Seurat or SingleCellExperiment object.

Let's try to classify cells in the test set:
```{r}
classified <- classify_cells(classify_obj = test_set, sce_assay = 'counts', 
                             cell_types = 'all', path_to_models = getwd())
```

Using the *classify_cells()* function, we have to indicate exactly the repository containing the database that the models has recently been saved to. In the previous section, we saved our new models to the current working directory. 

In the *classified* object, the classification process added new columns to the cell meta data, including the *predicted_cell_type* and *most_probable_cell_type* columns. 

If we use the full prediction to compare with actual plasma tag, we obtain this result: 
```{r}
# compare the prediction with actual cell tag
table(classified$predicted_cell_type, classified$plasma)
# plasma cell is child cell type of B cell
# so of course, all predicted plasma cells are predicted B cells 
```

When comparing the actual tag with the most probable prediction, we obtain: 
```{r}
# compare the prediction with actual cell tag
table(classified$most_probable_cell_type, classified$plasma)
```

The number of identified plasma cells is different in the *predicted_cell_type* slot and in the *most_probable_cell_type*. This is because the *predicted_cell_type* takes all predictions having the probabilities satisfying the corresponding probability thresholds. Meanwhile, the *most_probable_cell_type* takes only the cell type which gives highest prediction probability. 

To have all plasma cells specified as plasma cells, we can set the *ignore_ambiguous_result* to TRUE. This option will hide all ambiguous predictions where we have more than one possible cell type. In the parent-chid(ren) relationship of cell types, the more specified cell types/phenotypes will be reported.

```{r}
classified <- classify_cells(classify_obj = test_set, sce_assay = 'counts',
                             cell_types = 'all', path_to_models = getwd(),
                             ignore_ambiguous_result = TRUE)
table(classified$predicted_cell_type, classified$plasma)
```

## Session Info

```{r}
sessionInfo()
```